### 2015-03-19 doc/thesis/typedlua.tex
Removing the notation for fixed table types

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

@@

Typed Lua does not allow assignments to variables that hold a projection type,
unless the union has been discriminated down to a single value.
Unrestricted assignment to these variables would be unsound,
as it could break the dependency relation among the types in each tuple
that is part of the union.

@@

This example shows that we cannot write a function that is
guaranteed to return a number when we pass a number and
guaranteed to return a string when we pass a string.
Even though we could use intersection types to express the type
of this function as
\texttt{(number) -> (number) \string& (string) -> (string)},
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.

@@

\begin{figure}[!ht]
\textbf{Table types}\\
\dstart
\begin{align*}
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'} \; | \;
  \texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua table types}
\label{fig:tables}
\end{figure}

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.

Even though Typed Lua includes two different forms for annotating
a table, in this section we only discuss the form
$\texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}$.
We call this form \emph{closed} table type because it
represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In Sections \ref{sec:modules} and \ref{sec:oop} we will show that
we use the syntax $\texttt{`<\{'} \; [\textit{tabletypebody}] \; \texttt{`\}>'}$
to type modules and objects.
We call this form \emph{fixed} table type because it
represents tables with no keys that do not inhabit one of the table's
key types, and can have safe aliases.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.
In the next chapter we formalize and discuss in more detail the
different table types that our type system has.

@@

In this example, the type of \texttt{i} is \texttt{integer},
while the type of \texttt{x}, \texttt{y}, and \texttt{z} is
\texttt{string|nil}.
An inconvenient aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

@@

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the variable.
If we remove the annotation in the previous example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\texttt{\{1:string, 2:string, 3:string, 4:string, 5:string, 6:string, 7:string\}}.

When we use the syntax of records to type an array, the compiler
raises a warning when we try to access an index that is out of bounds.
In the previous example, both \texttt{days[8]} and \texttt{days[i]}
raise warnings, as both the literal type \texttt{8} and the base type
\texttt{integer} do not map to any value.

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
\emph{unique}, otherwise it can be either \emph{fixed} or
\emph{closed}.
In this section we discuss the cases that produce \emph{closed}
table types, while in Sections \ref{sec:modules} and \ref{sec:oop}
we will discuss the cases that produce \emph{fixed} table types
to describe the type of modules and classes.
Any alias to \emph{unique} variables makes them \emph{open}, as we use the
tag \emph{open} to keep track of \emph{unique} variables that are aliased.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

@@

We can create aliases to \emph{open} variables, but they are always
\emph{closed}.
Any mutation in the original reference and in the aliased reference
is not a problem, as the type of the original reference can only
add new fields.
For mutable fields this means that the type of the field cannot
change after it is added to the type of the table.
In our previous example, the second line changes the type of
\texttt{person} from \emph{unique} to \emph{open} and makes the type of
\texttt{bogus} \emph{closed}.

@@

@@

As an example, the following declaration defines the type \texttt{Person}
as an alias to the \emph{closed} table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = {"firstname":string,
                              "lastname":string}
\end{verbatim}

Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
for \emph{closed} table types that specify records, as writing table types can be unwieldy
when records get bigger and the types of record fields get more complicated.
Figure \ref{fig:interfaces} shows the concrete syntax of the
\texttt{interface} construct.

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{type}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring \emph{closed} table types that express records.
For this reason, it worth having two different constructs;
one specifically for records and another for more general types.

@@

We need an explicit type declaration on the return type because
the Typed Lua compiler cannot infer recursive types, though
it has subtyping rules to check that the inferred return type
matches the annotation.

@@

In this example, Typed Lua uses the type of the table constructor
to allow the programmer to build the type of the module.
The local variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we did not include it in the list of exported members. 
The \texttt{return} at the end of the source file gives the type of the module:
\begin{verbatim}
    <{ "deg":(number) -> (number),
       "rad":(number) -> (number),
       "pow":(number, number) -> (number) }>
\end{verbatim}

Typed Lua always fixes \emph{unique} and \emph{open} table types
that appear in a return statement.
Typed Lua has this behavior to ensure that modules and classes
always implement \emph{fixed} table types.
In the next section we will show that we build classes in a
similar way that we use to build modules, and we will also show
that classes should have \emph{fixed} table types to allow safe
inheritance.

@@

In this example, Typed Lua uses the refinement of table types to
allow the programmer to incrementally build the type of the module
\texttt{mymath}.
The variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we declared it as local to the module. 
The \texttt{return} at the end of the source file gives the
type of the module, which is the same type of the variable \texttt{mymath}.
This module has the same type of the module from the previous example.

@@

In Typed Lua, \texttt{require} is a primitive that statically type checks
a given module to infer the type of this module.
This means that the type of its input parameter must be a literal string,
as Typed Lua uses this literal string to find the source file that
implements a given module.
To statically type check a module, Typed Lua follows the same rules
that Lua follows to load a module.
Typed Lua first checks whether the module is already statically type checked.
When the module is not yet statically type checked, Typed Lua
statically type checks its source file, and the type returned is the
type of the module.
Typed Lua raises a compile-time warning when it cannot find the source
file of a given module.

@@

Lua provides minimal support for object-oriented programming.
The basic mechanism is the \texttt{:} syntactic sugar for method
definitions and method calls.
In the case of method definitions, the Lua compiler translates
\texttt{function obj:method(args) end} into an operation that
assigns a function to the field \texttt{method} in \texttt{obj}.
This function includes a first parameter named \texttt{self} plus
any other parameters.
In the case of method calls, the Lua compiler translates
\texttt{obj:method(args)} into an operation that first evaluates
\texttt{obj}, uses \texttt{method} to index \texttt{obj},
and then calls \texttt{obj.method} with the result of evaluating
\texttt{obj} as the first argument, followed by the result of
evaluating the argument list in the original expression.

Typed Lua uses table types along with the type
\texttt{self} to represent objects.
We use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

Typed Lua assigns the type \texttt{self} to the implicit parameter \texttt{self}.
The type \texttt{self} represents the type of the \emph{receiver}
in method definitions and method calls.
In this example, Typed Lua binds the type \texttt{self} to
the table type \texttt{\{"x":number, "y":number\}} inside \texttt{move}.
This example also uses the refinement of table types to build the
type of \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      "move":(self, number, number) -> () }
\end{verbatim}

@@

The first \texttt{setmetatable} rule appears in our previous example.
In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has type \texttt{self} then the expression also also has type \texttt{self}.
We will explain the other two rules while we explain how
Typed Lua type checks single inheritance.

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
Finally, the last \texttt{return} statement fixes the type of
\texttt{Shape} because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinment
of table types to type check single inheritance, as we will see in
this section.

A return statement that appears in the top-level also exports
a type alias to be used in type annotations.
For instance, our example exports the following type alias:
\begin{verbatim}
    typealias Shape =
      <{ "x":number, "y":number,
         const "new":(self, number, number) -> (self),
         const "move":(self, number, number) -> () }>
\end{verbatim}

The style of classes definition from the previous example
allow us to use \texttt{require} for creating prototype objects that
work as classes, along with an alias to the table type of the class.
This allows us to use the exported alias in type annotations,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

Typed Lua also includes a \texttt{class} construct that programmers
can use to document the type of the classes they export..
Figure \ref{fig:classes} shows the concrete syntax of the
\texttt{class} construct.

\begin{figure}[!ht]
\textbf{Classes}\\
\dstart
\begin{align*}
\textit{class} ::= & \;\; [\textbf{local}] \; \textbf{class} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua classes}
\label{fig:classes}
\end{figure}

As an example, we can use the following \texttt{class} declaration to
document the type \texttt{Shape} from the previous example:
\begin{verbatim}
    class Shape
      x, y:number
      const new:(number, number) => (self)
      const move:(number, number) => ()
    end
\end{verbatim}

We use a double arrow instead of a single arrow because it is
syntactic sugar for defining the type of methods,
as it includes an implicit first type \texttt{self}.
The double arrow can appear only inside the declaration of
classes and userdata, which we will introduce in the next section.

Typed Lua also uses \texttt{setmetatable} to simulate single inheritance.
We use the following example to discuss single inheritance in Typed Lua:
\begin{verbatim}
    local Shape = require "shape"
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0.0
    const function Circle:new (x:number, y:number, radius:value):self
      local c:self = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0.0
      return c
    end
    const function Circle:area ():number
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}

In this example, we use the other two \texttt{setmetatable} rules.
The second rule appears to trigger the refinement of table types,
as we need this rule to add new methods and also to override
existing methods in \texttt{Circle}.
In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has a \emph{fixed} table type then it is safe to
produce an equivalent \emph{open} table type, as \emph{fixed}
table types do not allow hiding fields.
The third rule appears to redefine the constructor \texttt{new}.
In a \texttt{setmetatable} expression
\texttt{setmetatable(e,\{\string_\string_index = id\})},
if \texttt{e} has a \emph{fixed} table type \texttt{t}, \texttt{id} has
type \texttt{self}, and the type bound to \texttt{self} is a subtype of
\texttt{t}, then the expression has type \texttt{self}.
Note how we can use this rule to call the constructor of \texttt{Shape}
inside the overridden constructor.
A limitation of this class system is that the overridden constructor
must be a subtype of the original constructor,
so the type of \texttt{radius} has to be quite permissive.

Like in the preivous example, the last return statement in the
top-level exports a \emph{fixed} table type, which we could
document through the following class declaration:
\begin{verbatim}
    class Circle
      x, y, radius:number
      const new:(number, number, value) => (self)
      const move:(number, number) => ()
      const area:() => (number)
    end
\end{verbatim}

Note that this class definition introduces a syntactical overhead
to manually include new and overriden methods.
This class definition is equivalent to the alias that exports the
following \emph{fixed} table type to the class \texttt{Circle}:
\begin{verbatim}
    <{ "x":number, "y":number, "radius":number,
       const "new":(self, number, number, value) -> (self),
       const "move":(self, number, number) -> (),
       const "radius":() -> (number) }>
\end{verbatim}

>>>>

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua always includes \texttt{nil*} in the end of the type
of an expression list that does not end in a varidic type.
This behavior preserves the semantics of Lua on replacing missing values,
and it is necessary when we omit optional parameters in a function call.
We will discuss optional parameters in the next section.

@@

Typed Lua does not allow assignments to variables that hold a projection type.
Unrestricted assignment to these variables would be unsound,
as it could break the dependency relation among the types in each tuple
that is part of the union.

@@

This example means that we cannot write a function that is
guaranteed to return a number when we pass a number and
guaranteed to return a string when we pass a string.
Even though we could use intersection types to express the type
of this function as
\texttt{(number) -> (number) \string& (string) -> (string)},
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.

@@

\begin{figure}[!ht]
\textbf{Table types}\\
\dstart
\begin{align*}
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua table types}
\label{fig:tables}
\end{figure}

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.
Later in this chapter we will show that we use the syntax of records
to type modules and objects.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.

@@

In this example, the type of \texttt{i} is \texttt{integer},
while the type of \texttt{x}, \texttt{y}, and \texttt{z} is
\texttt{string|nil}.
An inconvenient aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use the logical \texttt{or} operator or the \texttt{if}
statement to narrow the type of the elements they are accessing.

@@

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the variable.
If we remove the annotation in the previous example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\begin{verbatim}
    { 1:string, 2:string, 3:string, 4:string,
      5:string, 6:string, 7:string }
\end{verbatim}

When we use the syntax of records to type an array, the compiler
raises a warning when we try to access an index that is out of bounds.
In the previous example, the expressions \texttt{days[8]} and \texttt{days[i]}
would raise warnings if we had used the records syntax,
as both the literal type \texttt{8} and the base type
\texttt{integer} would not map to any value.

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
\emph{unique}, otherwise it is \emph{closed}.
Any alias to \emph{unique} variables makes them \emph{open}, as we use the
tag \emph{open} to keep track of \emph{unique} variables that are aliased.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
In Section \ref{sec:oop} we will show that we use the tag
\emph{fixed} to describe the type of classes.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

@@

We can create aliases to \emph{open} variables, but they are always \emph{closed}.
Any mutation in the original reference and in the aliased reference
is not a problem, as the type of the original reference can only
add new fields.
For mutable fields this means that the type of a field cannot
change after it is added to the type of a table.
In our previous example, the second line changes the type of
\texttt{person} from \emph{unique} to \emph{open} and makes the type of
\texttt{bogus} \emph{closed}.

@@

In Section \ref{sec:refinement} we present the formalization of the
rules that type check the refinement of table types.

@@

As an example, the following declaration defines the type \texttt{Person}
as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = {"firstname":string, "lastname":string}
\end{verbatim}

Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
for table types that specify records, as writing table types can be unwieldy
when records get bigger and the types of record fields get more complicated.
Figure \ref{fig:interfaces} shows the concrete syntax of the
\texttt{interface} construct.
In Section \ref{sec:oop} we will show that we can also use \emph{interface} to
document the type of objects, and that \emph{methodtype} is syntactic
sugar to express the type of methods.

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring table types that express records.
For this reason, it worth having two different constructs,
one specifically for records and another for more general types.

@@

We need an explicit type declaration in the return type because
the Typed Lua compiler cannot infer recursive types, though
it has subtyping rules to check that the inferred return type
matches the annotation.

@@

In this example, Typed Lua uses the type of the table constructor
to allow the programmer to build the type of the module.
The local variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we did not include it in the list of exported members. 
The \texttt{return} at the end of the source file gives the type of the module:
\begin{verbatim}
    { "deg":(number) -> (number),
      "rad":(number) -> (number),
      "pow":(number, number) -> (number) }
\end{verbatim}

Typed Lua always fixes \emph{unique} and \emph{open} table types
that appear in a top-level return statement.
This means that modules have \emph{fixed} table types in Typed Lua.
Typed Lua has this behavior to ensure that classes always implement
\emph{fixed} table types.
In the next section we will show that we build classes in a
similar way that we use to build modules, and we will also show
that classes should have \emph{fixed} table types to allow safe
single inheritance.

@@

In this example, Typed Lua uses the refinement of table types to
incrementally build the type of the module \texttt{mymath}.
The variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we declared it as local to the module. 
The \texttt{return} at the end of the source file gives the
type of the module, which is the same type of the variable \texttt{mymath}.
This module has the same type of the module from the previous example.

@@

In Typed Lua, \texttt{require} is a primitive that statically type checks
a given module to infer the type of this module.
This means that the type of its input parameter must be a literal string,
as Typed Lua uses this literal string to find the source file that
implements the given module.
To statically type check a module, Typed Lua follows the same rules
that Lua follows to load a module.
Typed Lua first checks whether the module is already statically type checked.
When the module is not yet statically type checked, Typed Lua
statically type checks its source file, and the type returned is the
type of the module.
Typed Lua raises a compile-time warning when it cannot find the source
file of a given module.

@@

Lua provides minimal support for object-oriented programming.
The basic mechanism is the \texttt{:} syntactic sugar for method
definitions and method calls.
In the case of method definitions, the Lua compiler translates
\texttt{function obj:method(args) end} into an operation that
assigns a function to the field \texttt{method} in \texttt{obj}.
This function includes a first parameter named \texttt{self} plus
any other parameters.
In the case of method calls, the Lua compiler translates
\texttt{obj:method(args)} into an operation that evaluates
\texttt{obj}, uses \texttt{method} to index \texttt{obj},
and then calls \texttt{obj.method} with the result of evaluating
\texttt{obj} as the first argument, followed by the result of
evaluating the argument list in the original expression.

Typed Lua uses \emph{closed} table types along with the type
\texttt{self} to represent objects.
We use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

Typed Lua assigns the type \texttt{self} to the implicit parameter \texttt{self}.
The type \texttt{self} represents the type of the \emph{receiver}
in method definitions and method calls.
In this example, Typed Lua binds the type \texttt{self} to
the \emph{closed} table type \texttt{\{"x":number, "y":number\}} inside \texttt{move}.
This example also uses the refinement of table types to build the
type of \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      "move":(self, number, number) -> () }
\end{verbatim}

@@

The first \texttt{setmetatable} rule appears in our previous example.
In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has type \texttt{self} then the expression also has type \texttt{self}.
We will explain the other two rules while we explain how
Typed Lua type checks single inheritance.

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
Finally, the top-level \texttt{return} statement fixes the type of
\texttt{Shape}, because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinment
of table types to type check single inheritance, as we will see in
this section.

A return statement that appears in the top-level also exports an
interface which we can use in type annotations.
For instance, our previous example exports the following interface:
\begin{verbatim}
    interface Shape
      x, y:number
      const new:(number, number) => (self)
      const move:(number, number) => ()
    end
\end{verbatim}

We use a double arrow instead of a single arrow because it is
syntactic sugar for defining the type of methods,
as it includes an implicit first type \texttt{self}.
The double arrow can appear only inside the declaration of
interfaces and userdata, which we will introduce in the next section.

The style of classes definition from the previous example
allows us to use \texttt{require} for creating prototype objects that
work as classes, along with an alias to the object type.
This allows us to use the exported alias in type annotations,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

Note that \texttt{Shape} has a \emph{fixed} table type, because it
describes the type of the class \texttt{Shape}, but \texttt{shape1}
and \texttt{shape2} have \emph{closed} table types, because they
describe the type of objects from the class \texttt{Shape}.
Now, we will see that we need \emph{fixed} table types to allow
\texttt{setmetatable} to simulate single inheritance.
We use the following example to discuss single inheritance in Typed Lua:
\begin{verbatim}
    local Shape = require "shape"
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0.0
    const function Circle:new (x:number, y:number, radius:value):self
      local c:self = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0.0
      return c
    end
    const function Circle:area ():number
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}

In this example, we use the other two \texttt{setmetatable} rules.
The second rule appears to trigger the refinement of table types,
as we need this rule to add new methods and also to override
existing methods in \texttt{Circle}.
The third rule appears to redefine the constructor \texttt{new}.

In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has a \emph{fixed} table type then it is safe to
produce an equivalent \emph{open} table type, as \emph{fixed}
table types do not allow hiding fields.
In our example, it is safe to use the type of \texttt{Shape}
to initialize \texttt{Circle}, openning the type of \texttt{Circle}
to allow the refinement of table types to build the type of
the class \texttt{Circle}.

In a \texttt{setmetatable} expression
\texttt{setmetatable(e:(el),\{\string_\string_index = id\})},
if the expression \texttt{e:(el)} has a \emph{closed} table type \texttt{t},
\texttt{id} has type \texttt{self}, and the type bound to \texttt{self}
is a subtype of \texttt{t}, then the expression has type \texttt{self}.
Note how we can use this rule to call the constructor of \texttt{Shape}
inside the overridden constructor.
A limitation of this class system is that the overridden constructor
must be a subtype of the original constructor,
so the type of \texttt{radius} has to be quite permissive.
Also note that this is the only form of refinement that allows
changing the type of a table field, if the new type is a subtype
of the previous type.

Like in the example that we introduced the class \texttt{Shape},
in the example that we introduced the class \texttt{Circle},
the top-level return statement exports an
interface which we can use in type annotations:
\begin{verbatim}
    interface Circle
      x, y, radius:number
      const new:(number, number, value) => (self)
      const move:(number, number) => ()
      const area:() => (number)
    end
\end{verbatim}

Even though the class \texttt{Circle} is not a subtype of the
class \texttt{Shape}, because \emph{fixed} table types do not
have width subtyping, the objects that have the exported type
\texttt{Circle} are subtypes of the objects that have the
exported type \texttt{Shape}, because \emph{closed} table types
have width subtyping.
We discuss \emph{fixed} and \emph{closed} table types in more
detail in the next chapter.

We can use both exported aliases in type annotations, as we
show in the following example:
\begin{verbatim}
    local Circle = require "circle"
    local circle1 = Circle:new(0, 5, 10)
    local circle2:Circle = Circle:new(10, 10, 10)
    local circle3:Shape = circle1
    local circle4:Shape = circle2
    print(circle2:area())             --> 314.15926535898
    print(circle3:area())             --> compile-time warning
\end{verbatim}
